---
title: "Gak 1 Temperature"
author: "Ben Williams"
date: "June 10, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message=FALSE)
```

Code to produce temperature figure from the GAK 1 line.

```{r load}
library(tidyverse)
library(extrafont)
loadfonts(device="pdf") # note that a Mac is different I'm on a windows machine - so you may have to turn this off
theme_set(theme_bw(base_size=12, base_family='Times')+ 
             theme(panel.grid.major = element_blank(), 
                   panel.grid.minor = element_blank()))
```

Get the data 

```{r data}
gak1 <- scan("http://www.ims.uaf.edu/gak1/data/TimeSeries/gak1.dat", 
             what = list("", 0,0,0,0,0,0,0),skip=3)
```

Structure the lists as a dataframe and name them

```{r}
gak1 <- data.frame(gak1)
colnames(gak1) <- c("cruise","station","dec.year","depth","temp","sal","sigma.T","delta.D")
```

Structure the data
```{r}
gak1 %>% 
	mutate(year = trunc(dec.year),
			 Year = factor(year),
			 Depth=factor(depth),
			 Month = cut(dec.year-year, seq(0,1,length=13), labels = c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")),
			 month=as.numeric(Month)) -> gak1

```

EDA
```{r}
ggplot(gak1, aes(year, temp, fill=Depth, color=Depth))+
	geom_point()

ggplot(gak1, aes(year, temp, fill=Depth, color=Depth))+
	stat_summary(fun.y=mean, geom='line')

ggplot(gak1, aes(year, temp))+
	stat_summary(fun.y=mean, geom='line') + stat_smooth()
```

Unequal sampling of temps
```{r}
table(gak1$year,gak1$month)
```

```{r}
gak1 %>% 
	ggplot(aes(year)) + geom_density(alpha=.4, fill=4)
gak1 %>% 
	ggplot(aes(depth)) + geom_histogram(alpha=.4, fill=4, bins=12, color='white') 
```



Figure 1. Using Loess

```{r}
# This is the .tiff figure format for Plos One
#tiff(filename = "Fig1.tiff", width = 5.2, height = 3.2, units = "in", res=600, compression="lzw")
ggplot(gak1, aes(year, temp, fill=Depth, color=Depth))+
	stat_smooth(alpha=.10) + 
	scale_x_continuous(breaks=1970:2017,
							 labels=c(1970,rep("",4),1975,rep("",4),1980,rep("",4),1985,rep("",4),
							 			1990,rep("",4),1995,rep("",4),2000,rep("",4),2005,rep("",4),
							 			2010,rep("",4),2015,rep("",2))) + 
	scale_y_continuous(breaks=seq(3.5,9.5,1)) +
	scale_color_discrete(name='Depth (m)') + 
	scale_fill_discrete(name='Depth (m)') + 
	theme(legend.position=c(.75, .18), legend.background = element_rect(fill="transparent")) + 
	guides(fill=guide_legend(ncol=4), color=guide_legend(ncol=4)) +
	xlab('Year') + ylab(expression('Temperature ('*~degree*C*')'))
	
#dev.off()
```


Figure 2. Using gam - control the knots by changing *k*

```{r}
# This is the .tiff figure format for Plos One
#tiff(filename = "Fig2.tiff", width = 5.2, height = 3.2, units = "in", res=600, compression="lzw")
ggplot(gak1, aes(year, temp, fill=Depth, color=Depth))+
	stat_smooth(alpha=.10, method='gam', formula=y~s(x, k=10)) + 
	scale_x_continuous(breaks=1970:2017,
							 labels=c(1970,rep("",4),1975,rep("",4),1980,rep("",4),1985,rep("",4),
							 			1990,rep("",4),1995,rep("",4),2000,rep("",4),2005,rep("",4),
							 			2010,rep("",4),2015,rep("",2))) + 
	scale_y_continuous(breaks=seq(3.5,9.5,1)) +
	scale_color_discrete(name='Depth (m)') + 
	scale_fill_discrete(name='Depth (m)') + 
	theme(legend.position=c(.75, .18), legend.background = element_rect(fill="transparent")) + 
	guides(fill=guide_legend(ncol=4), color=guide_legend(ncol=4)) +
	xlab('Year') + ylab(expression('Temperature ('*~degree*C*')'))
	
#dev.off()
```

Figure 3. This is a bootstrap mean and 95% CI by year 

```{r}
# This is the .tiff figure format for Plos One
#tiff(filename = "Fig3.tiff", width = 5.2, height = 3.2, units = "in", res=600, compression="lzw")
ggplot(gak1, aes(year, temp, fill=Depth, color=Depth))+
	stat_summary(fun.data=mean_cl_boot, geom='smooth') + 
	scale_x_continuous(breaks=1970:2017,
							 labels=c(1970,rep("",4),1975,rep("",4),1980,rep("",4),1985,rep("",4),
							 			1990,rep("",4),1995,rep("",4),2000,rep("",4),2005,rep("",4),
							 			2010,rep("",4),2015,rep("",2))) + 
	scale_y_continuous(breaks=seq(3.5,9.5,1)) +
	scale_color_discrete(name='Depth (m)') + 
	scale_fill_discrete(name='Depth (m)') + 
	theme(legend.position=c(.75, .18), legend.background = element_rect(fill="transparent")) + 
	guides(fill=guide_legend(ncol=4), color=guide_legend(ncol=4)) +
	xlab('Year') + ylab(expression('Temperature ('*~degree*C*')'))
	
#dev.off()
```

Figure 4. This is a bootstrap mean and 95% CI by year for all depths

```{r}
# This is the .tiff figure format for Plos One
#tiff(filename = "Fig4.tiff", width = 5.2, height = 3.2, units = "in", res=600, compression="lzw")
gak1.deep=gak1[gak1$depth>100,]

ggplot(gak1.deep, aes(year, temp))+
	stat_summary(fun.data=mean_cl_boot, geom='smooth') + 
	scale_x_continuous(breaks=1970:2017,
							 labels=c(1970,rep("",4),1975,rep("",4),1980,rep("",4),1985,rep("",4),
							 			1990,rep("",4),1995,rep("",4),2000,rep("",4),2005,rep("",4),
							 			2010,rep("",4),2015,rep("",2))) + 
	scale_y_continuous(breaks=seq(3.5,9.5,1)) +
	xlab('Year') + ylab(expression('Temperature ('*~degree*C*')'))
	
#dev.off()
```


Figure 5. This is gam smoothed and 95% CI by year

```{r}
# This is the .tiff figure format for Plos One
#tiff(filename = "Fig5.tiff", width = 5.2, height = 3.2, units = "in", res=600, compression="lzw")
ggplot(gak1, aes(year, temp))+
	stat_smooth(method='gam', formula=y~s(x, k=10)) + 
	scale_x_continuous(breaks=1970:2017,
							 labels=c(1970,rep("",4),1975,rep("",4),1980,rep("",4),1985,rep("",4),
							 			1990,rep("",4),1995,rep("",4),2000,rep("",4),2005,rep("",4),
							 			2010,rep("",4),2015,rep("",2))) + 
	scale_y_continuous(breaks=seq(3.5,9.5,1)) +
	xlab('Year') + ylab(expression('Temperature ('*~degree*C*')'))
	
#dev.off()
```