---
title: "Fit DFA to Ichthyo Abundance"
author: "Kristin Marshall"
date: "August 23, 2016"
output:
  html_document: default
  word_document: default
---
This document contains the code and plots for fitting a DFA to the ichthyo. abundance data. The sections below step througg the model fitting and model selection procedure, show how I fit various covariates, and plots DFA fits and common trends. I also included a cluster analysis that groups species based on their loadings values, and a plot that summarizes those results.  I call out in the text 3 potential figures to include in the final manuscript. I'm imagining Figure 1 will be a map of the survey area, and Figures 2-4 are labeled below.
```{r, include=FALSE}
library(MARSS)
library(xtable)
library(calibrate)
library(ggplot2)
library(ggrepel)
library(rioja)
library(ggsidekick)
#setwd('/users/kristinmarshall/dropbox/Ichthyo_IndexStandardization/')

getwd()
abund=read.csv('../data/DensityForDFA.csv', header=T)

spnames=rev(names(abund))
names(abund)=spnames

##replace years without data with NAs-- 1984, 1986, 2012.  survey starts in 1981 so years 4,6,32 are the ones missing 
abund[4,]=NA
abund[6,]=NA
abund[32,]=NA

abund.cut=abund

aborder= order(colSums(abund.cut, na.rm=T), decreasing=T)
abund.order=abund.cut[,aborder]

# transpose data so time goes across columns
dat.abund = t((abund.order))
#dat.abund=dat.abund[-1,] #removes teh year row
# get number of time series
N.ts = dim(dat.abund)[1]
# get length of time series
TT = dim(dat.abund)[2] 
#years for plotting
years.plot=seq(1981,2013)

Sigma = sqrt(apply(dat.abund, 1, var, na.rm=TRUE))
y.bar = apply(dat.abund, 1, mean, na.rm=TRUE)
dat.z = (dat.abund - y.bar) * (1/Sigma)
rownames(dat.z) = rownames(dat.abund)

spp = rownames(dat.abund)

#pdf("RawAbund.pdf")
par(mfrow=c(4,5), mar=c(0,0,0,0), oma=c(4,4,1,1))
par(mgp = c(.6, .6, 0))
par(tcl=.2)
for(i in 1:length(spp)){
  plot(years.plot,dat.z[i,],xlab="",ylab="Index", bty="L", pch=16, col="blue", type="b", ylim=c(min(dat.z, na.rm=T),max(dat.z, na.rm=T)), axes=F)
  
  if(i %in% c(16:20,36:40)) axis(1,at=seq(1990,2010, by=10), cex.axis=0.8)
  if(i %in% c(1,6,11,16, 21, 26, 31,36)) axis(2, at=seq(-2,5, by=1), cex.axis=0.8)
  box()
  mtext(spp[i], side=3, line=-1, cex=0.5)
}


mtext("Year", side=1, outer=TRUE, line=2, cex=0.7)
mtext("Standardized abundance", side=2, outer=TRUE, line=2, cex=0.7)

#dev.off()
```

## Fit the DFA models
Next, we fit all of the potential DFA models without covariates.  This included 1 to 4 common trends, and 3 different variance structures (diagonal and equal, diagonal and unequal, and equalvarcov).  A previous iteraion suggested the uncontrained variance structure wouldn't converge.  Given we're working with 40 time series, that's not surprising because it would require estimating 1600 parameters, and we only have 1200 data points.

```{r, eval=FALSE}

cntl.list = list(minit=200, maxit=3000, allow.degen=FALSE)
# set up forms of R matrices
levels.R = c("diagonal and equal",
             "diagonal and unequal",
             "equalvarcov")

model.data=data.frame()
for (j in 2:3){
  for (i in 1:3){
    
    model.list = list(m=i, R=levels.R[j])
    kemz = MARSS(dat.abund, model=model.list, control=cntl.list,
                 z.score=TRUE, form="dfa")
    
    model.data = rbind(model.data, 
                       data.frame(R=levels.R[j],
                                  m=i,
                                  logLik=kemz$logLik,
                                  K=kemz$num.params,
                                  AICc=kemz$AICc,
                                  converg=kemz$convergence,
                                  stringsAsFactors=FALSE))
    
    assign(paste("kemz", i, levels.R[j], sep="."), kemz)
    
    
  }
}

###################################################
### makemodeltable
###################################################
# calculate delta-AICc
model.data$delta.AICc = model.data$AICc - min(model.data$AICc)
# calculate Akaike weights
wt = exp(-0.5*model.data$delta.AICc)
model.data$Ak.wt = wt/sum(wt)
# sort results
model.tbl = model.data[order(model.data$AICc),-4]
# drop AICc from table
# calculate cumulative wts
model.tbl$Ak.wt.cum = cumsum(model.tbl$Ak.wt)

print(model.tbl)

write.csv(model.tbl,'../data/AbundDFA_AIC.csv')
```
We output the model table here to show that the best-fit model is a two trend model with a covariance matrix with a single parameter on the diagonal, and a second parameter on the diagonal (equalvarcov). 
```{r, echo=FALSE}
model.tbl=read.csv('../data/AbundDFA_AIC.csv', header=T)
print(model.tbl)
```
##Fit DFA models with covariates
The next step is to see if we can improve the model fit by adding explanatory covariates.  We looked at three types of covariates: environmental indices, an index for the EVOS spill year, spawning stock biomasses of pollock, pcod, and arrowtooth. The take-home was that none of these covariates improved the AICc over the two trend model.  I output the code below.

###Environmental covariates
First, let's look at the environmental covariates.  We looked for relationships with SST, ENSO, NPGO, PDO, NPI, and Upwelling.
```{r, eval=FALSE}
temp=read.csv('../data/SSTcovar.csv', header=T)
climate=read.csv('../data/ClimateCovars.csv',header=T)
asst=temp$asst[1:33]
wsst=temp$wsst[1:33]
enso=climate$ENSO[7:39]
npgo=climate$NPGO[7:39]
npi=climate$NPI[7:39]
pdo=climate$PDO[7:39]
up=climate$Upwelling[7:39]

model.list = list(m=2, R="equalvarcov")
cntl.list=list(minit=200, maxit=5000, allow.degen=FALSE)
kemz.no= MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa")
kemz.2.asst = MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa", covariates=asst)
kemz.2.wsst=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=wsst)
kemz.2.enso=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=enso)
kemz.2.npgo=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=npgo)
kemz.2.npi=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=npi)
kemz.2.pdo=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=pdo)
kemz.2.up=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=up)

env.out=cbind(model=c("no covars", "AnnTemp", "WinTemp","ENSO","NPGO","NPI","PDO","Upwelling"),
            AICc=round(c(kemz.no$AICc, kemz.2.asst$AICc, kemz.2.wsst$AICc, kemz.2.enso$AICc, 
                         kemz.2.npgo$AICc, kemz.2.npi$AICc, kemz.2.pdo$AICc,kemz.2.up$AICc)))
print(env.out, quote=F)

# model     AICc
#[1,] no covars 3261
#[2,] AnnTemp   3354
#[3,] WinTemp   3355
#[4,] ENSO      3313
#[5,] NPGO      3305
#[6,] NPI       3354
#[7,] PDO       3301
#[8,] Upwelling 3312

```
From this table, we conclude that none of the climate cocariates improved the model fit over a 2 trend model. 

Similarly, we might want to know if adding these covariates to a one trend model improves the fit.
```{r, eval=FALSE}
model.list = list(m=1, R="equalvarcov")
cntl.list=list(minit=200, maxit=5000, allow.degen=FALSE)
kemz.1.no= MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa")
kemz.1.asst = MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa", covariates=asst)
kemz.1.wsst=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=wsst)
kemz.1.enso=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=enso)
kemz.1.npgo=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=npgo)
kemz.1.npi=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=npi)
kemz.1.pdo=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=pdo)
kemz.1.up=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=up)

env.1.out=cbind(model=c("no covars", "AnnTemp", "WinTemp","ENSO","NPGO","NPI","PDO","Upwelling"),
            AICc=round(c(kemz.1.no$AICc, kemz.1.asst$AICc, kemz.1.wsst$AICc, kemz.1.enso$AICc, 
                         kemz.1.npgo$AICc, kemz.1.npi$AICc, kemz.1.pdo$AICc,kemz.1.up$AICc)))

print(env.1.out, quote=FALSE)

#model     AICc
#no covars 3280
#[2,] AnnTemp   3369
# [3,] WinTemp   3369
# [4,] ENSO      3328
# [5,] NPGO      3321
# [6,] NPI       3369
# [7,] PDO       3321
# [8,] Upwelling 3328
```
We conclude that including environmental covariates do not improve the one trend model, either. 

###Fit SSB covariates
Next, we try it for pollock, pcod, and arrowthooth, for both the 1 trend and 2 trend models.
```{r, eval=FALSE}
SSB=read.csv('../data/ssbout.csv', header=T)
#ind=which(SSB$Year >1980 & SSB$Year<2014)
poll=SSB$V2/10^5
pcod=SSB$V3[ind]/10^5
arrtooth=SSB$V5[ind]/10^6

model.list = list(m=2, R="equalvarcov")
cntl.list=list(minit=200, maxit=5000, allow.degen=FALSE)
kemz.2.no= MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa")
kemz.2.poll = MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa", covariates=poll)
kemz.2.pcod=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=pcod)
kemz.2.arr=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=arrtooth)

ssb2.out=cbind(model=c("2 trend", "2.Pollock", "2.Pcod","2.Arrowtooth"),
            AICc=round(c(kemz.2.no$AICc, kemz.2.poll$AICc, kemz.2.pcod$AICc, kemz.2.arr$AICc)))


model.list = list(m=1, R="equalvarcov")
cntl.list=list(minit=100, maxit=5000, allow.degen=FALSE)
kemz.1.no= MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa")
kemz.1.poll = MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa", covariates=poll)
kemz.1.pcod=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=pcod)
kemz.1.arr=MARSS(dat.abund, model=model.list, control=cntl.list,
                z.score=TRUE, form="dfa", covariates=arrtooth)

ssb1.out=cbind(model=c("1 trend", "1.Pollock", "1.Pcod","1.Arrowtooth"),
            AICc=round(c(kemz.1.no$AICc, kemz.1.poll$AICc, kemz.1.pcod$AICc, kemz.1.arr$AICc)))


print(ssb1.out, quote=FALSE)
# model        AICc
# [1,] 1 trend      3280
# [2,] 1.Pollock    3362
# [3,] 1.Pcod       3361
# [4,] 1.Arrowtooth 3367
print(ssb2.out, quote=FALSE)
```
None of the SSB indices helped model fit, and many didn't converge with the 2 trend model.

###Fit oil spill covariate
Last, we want to know if there is support for the year of the spill being anamolous. So, we creates a covariate for spill year, which is 0 in every year except 1989, in that year it's -1. Then we ask if that improves the model fit.

```{r, eval=FALSE}
spill=rep(0,33)
spill[9]=-1

model.list = list(m=1, R="equalvarcov")
cntl.list=list(minit=200, maxit=5000, allow.degen=FALSE)
kemz.1.no= MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa")
kemz.1.oil = MARSS(dat.abund, model=model.list, control=cntl.list,
                  z.score=TRUE, form="dfa", covariates=spill)
model.list = list(m=2, R="diagonal and equal")
kemz.2.no= MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa")
kemz.2.oil = MARSS(dat.abund, model=model.list, control=cntl.list,
                  z.score=TRUE, form="dfa", covariates=spill)

print(cbind(model=c("1 trend", "1 trend + oil", "2 trend","2 trend + oil"),
            AICc=round(c(kemz.1.no$AICc, kemz.1.oil$AICc, kemz.2.no$AICc, kemz.2.oil$AICc))), quote=FALSE)
```

We saw no evidence of an oil spill effect using this approach.  The two trend model with equalvarcov is still the best.

##Best fit DFA
Let's plot the best fit model for each species and calculate the percent variance explained. This is **Appendix C** in the paper.

```{r, include=FALSE, eval=F, message=FALSE}
##enter best model parameters here
model.list = list(m=2, R="equalvarcov")
cntl.list=list(minit=200, maxit=5000, allow.degen=FALSE)
kemz.2 = MARSS(dat.abund, model=model.list, control=cntl.list,
               z.score=TRUE, form="dfa")

saveRDS(kemz.2,"../data/best_fit_DFA.rds")
```
```{r, echo=FALSE}
###################################################
### plotfits
###################################################
kemz.2=readRDS("../data/best_fit_DFA.rds")
fit = kemz.2
spp = rownames(dat.z)
#pdf("NEWAbundDFAFits.pdf")
par(mfrow=c(4,5), mar=c(0,0,0,0), oma=c(4,4,1,1))
par(mgp = c(.6, .6, 0))
par(tcl=.2)
for(i in 1:length(spp)){
  plot(years.plot,dat.z[i,],xlab="",ylab="Index", bty="L", pch=16, col="blue",
       ylim=c(min(dat.z, na.rm=T),max(dat.z, na.rm=T)), xlim=c(1980,2014),axes=F)
  par.mat=coef(fit,type="matrix")
  lines(years.plot,as.vector(par.mat$Z[i,,drop=FALSE]%*%fit$states+par.mat$A[i,]), lwd=2)
  if(i %in% c(16:20,36:40)) axis(1,at=seq(1980,2010, by=5), cex.axis=0.8)
  if(i %in% c(1,6,11,16, 21, 26, 31,36)) axis(2, at=seq(-2,5, by=1), cex.axis=0.8)
  box()
  mtext(spp[i], side=3, line=-1, cex=0.5)
}
mtext("Year", side=1, outer=TRUE, line=2, cex=0.7)
mtext("Standardized abundance", side=2, outer=TRUE, line=2, cex=0.7)
#dev.off()
```

##Indices of Abundance
Here are the plots of the standardized index of abundance by species, **Appendix B** in the paper.
```{r}
#pdf("plots/AppB.RawAbund.pdf")
par(mfrow=c(4,5), mar=c(0,0,0,0), oma=c(4,4,1,1))
par(mgp = c(.6, .6, 0))
par(tcl=.2)
for(i in 1:length(spp)){
  plot(years.plot,dat.z[i,],xlab="",ylab="Index", bty="L", pch=16, col="blue",
       ylim=c(min(dat.z, na.rm=T),max(dat.z, na.rm=T)), xlim=c(1980,2014),axes=F)
  par.mat=coef(fit,type="matrix")
  lines(years.plot,as.vector(par.mat$Z[i,,drop=FALSE]%*%fit$states+par.mat$A[i,]), lwd=2)
  if(i %in% c(16:20,36:40)) axis(1,at=seq(1980,2010, by=5), cex.axis=0.8)
  if(i %in% c(1,6,11,16, 21, 26, 31,36)) axis(2, at=seq(-2,5, by=1), cex.axis=0.8)
  box()
  mtext(spp[i], side=3, line=-1, cex=0.5)
}
mtext("Year", side=1, outer=TRUE, line=2, cex=0.7)
mtext("Standardized abundance", side=2, outer=TRUE, line=2, cex=0.7)

```


In addition to looking at the fits, we want to look at the common **trends** and **loadings**.  Before we can do that, we need to rotate the trends and loadings to make them as orthogonal as possible.

```{r}
best.fit=kemz.2
###################################################
### varimax
###################################################
# get the inverse of the rotation matrix
H.inv = varimax(coef(best.fit, type="matrix")$Z)$rotmat


###################################################
### rotations
###################################################
Z.no.rot=coef(best.fit, type="matrix")$Z
# rotate factor loadings
Z.rot = coef(best.fit, type="matrix")$Z %*% H.inv   

##use this if only a 1-trend model
#Z.rot=coef(best.fit, type="matrix")$Z


# rotate trends
trends.rot = solve(H.inv) %*% best.fit$states

##use this if only a 1-trend model
#trends.rot=best.fit$states
trends.no.rot=best.fit$states
```
Now, we can plot the rotated trends. The trend plot is **Figure 3**.
```{r, echo=FALSE}
###################################################
### plotfacloadings
##################################################
sp.names=read.csv("../data/Abund.names.csv", header=T)

# minZ = 0.2
# ylims = c(-1.1*max(abs(Z.rot)), 1.1*max(abs(Z.rot)))
# #pdf("AbundDFAloadings2trend.pdf")
# par(mfrow=c(1,2), mar=c(1,3,0,0), oma=c(1,1,1,1), xpd=T)
# par(mgp = c(.6, .6, 0))
# barplot(Z.rot[,1], horiz=TRUE, xlim=c(-0.9,0.8), xlab='', cex.axis=0.6)
# box()
# text(x=-0.95,y=seq(from=1,by=1.2, length.out=40), lab=sp.names$ShortName, xpd=T, adj=1, cex=0.5)
# 
# 
# par(mar=c(1,0,0,3))
# barplot(Z.rot[,2], horiz=TRUE, xlim=c(-0.9,0.8), xlab='', cex.axis=0.6)
# box()

###################################################
### plottrends
###################################################
# get ts of trends
ts.trends = t(trends.rot)
plab=c('a','b')
#pdf("AbundDFA2trends.pdf")
par(mfrow=c(ceiling(dim(ts.trends)[2]/2),2), mar=c(3,2,1.5,0.5), oma=c(0.4,1,1,1))
# loop over each trend
for(i in 1:dim(ts.trends)[2]) {
  # set up plot area
  plot(seq(1981,2013),ts.trends[,i],
       ylim=c(-1.1,1.1)*max(abs(ts.trends)), 
       type="n", lwd=2, bty="L", 
       xlab="", ylab="",  yaxt="n")
  # draw zero-line
  abline(h=0, col="grey80")
   abline(v=seq(1985,2010,by=5), lty=1, col='grey80')
  # plot trend line
  par(new=TRUE)
  plot(seq(1981,2013),ts.trends[,i],
       ylim=c(-1.1,1.1)*max(abs(ts.trends)), 
       type="o", lwd=2, bty="L", pch=16,
       xlab="", ylab="", xaxt="n")
 
  # add panel labels
  #mtext(paste("Trend",i,sep=" "), side=3, line=0.5)
  mtext(paste(plab[i],')', sep=""), side=3, line=0, adj=-0.1,cex=1, font=2)
  #mtext('b)', side=3, line=0, adj=-0.2, cex=1)
  #axis(1,12*(0:dim(dat.abund)[2])+1,1980+0:dim(dat.abund)[2])
} # end i loop (trends)

#dev.off()
```

We can also calculate the proportion of variance explained by the full model, using sum of squared errrors.
```{r}
##rsquared
pred = Z.no.rot %*% trends.no.rot
SSres = sum((pred - dat.z) ^ 2, na.rm=T)
SStot = sum((mean(dat.z, na.rm=T) - dat.z) ^ 2, na.rm=T)
R2 = 1 - SSres / SStot
print(R2)
```

```{r, include=FALSE}
#write.csv(ts.trends,"Abund.trends2.csv")
#load=cbind(spp,Trend1=Z.rot[,1],Trend2=Z.rot[,2])
#write.csv(load, "Abund.loadings.csv", row.names=F)

#ts.trends=read.csv("../data/Abund.trends2.csv",header=T)[,2:3]
#Z.rot=read.csv("../data/Abund.loadings.csv", header=T)[,2:3]

#par(mfrow=c(1,2))
#plot(seq(1981,2013),trends.rot, type='l')
#plot(seq(1981,2013),ts.trends[,1], type='l')


```

Now, let's look at how the individual species group together, based on their loadings on the common trends.  Here is a cluster analysis based on euclidian distances.  The cluster dendrogram could appear in the supplementary material, but I probably wouldn't put it in the main text.  

```{r}
##cluster diagrams of loadings

di <- dist(Z.rot[,1:2], method="euclidean",diag = TRUE, upper = TRUE)



#x2=chclust(di)
#plot(x2)
#bstick(x2)


x=hclust(di, method = "complete", members=NULL)
plot(x)
y=rect.hclust(x, k=5)

## S3 method for class 'hclust'
#pdf("AbundClust2trend.pdf")
# par(mfrow=c(1,1),mar=c(1,4,0,0), oma=c(0,0,0,0))
# plot(x2, labels = sp.names$LongName, hang = 0.1,
#      axes = TRUE, frame.plot = FALSE, ann = TRUE,
#      main = "",
#      sub = NULL, xlab = "", ylab = "Height")
```

Breaking the dendrogram into 4 groups, we can then make a bivariate plot of the loadings, showing the four different groups using different colored text. This plot is **Figure 4**.
```{r, echo=FALSE}
#sp.names$groups=NA
#sp.names$group[x$order[1:14]]=1
#sp.names$group[x$order[15:25]]=2
#sp.names$group[x$order[26:33]]=3
#sp.names$group[x$order[34:40]]=4


set.seed(30)

no.plot=which(abs(Z.rot[,1])<0.2 & abs(Z.rot[,2])<0.2)

no.lab=c(no.plot,1,4,5,8,9,14,31,32,33, 36)
##where to plot labels
#gadus moves down, bathyagonus moves up, ammodytes moves up, gadus moves up, platichtys moves down, icelinus moves up, protomuctopum moves up
#lab.Zrotx=Z.rot[-no.plot,1] * 1.1

#lab.Zroty=Z.rot[-no.plot,2] * 1.1

row.names(Z.rot)=sp.names$spp
row.names(Z.rot)[1]="Gadus.chalcogrammus"
Z.rot.plot=Z.rot[-no.plot,]

Z.rot.plot$Group=5
Z.rot.plot$Group[Z.rot.plot$Trend1>.2 & abs(Z.rot.plot$Trend2)<.2]=1
Z.rot.plot$Group[Z.rot.plot$Trend1<(-.2) & abs(Z.rot.plot$Trend2)<.2]=2
Z.rot.plot$Group[abs(Z.rot.plot$Trend1)<.2 & Z.rot.plot$Trend2>.2]=3
Z.rot.plot$Group[abs(Z.rot.plot$Trend1)<.2 & Z.rot.plot$Trend2<(-.2)]=4
Z.rot.plot$Group=factor(Z.rot.plot$Group, labels=c('Strong Positive Tr 1','Strong Negative Tr 1.','Strong Positive Tr 2','Strong Neg Tr 2.','Others'))



#pdf("LoadingsPlotv2.pdf", width=6, height=6)
ggplot(Z.rot.plot) +
 #geom_vline(xintercept=0, colour="grey80", size=2 ) +
 #geom_hline(yintercept = 0, colour="grey80", size=2) +
 geom_point(aes(Trend1,Trend2, color=Group)) +
 geom_text_repel(aes(Trend1, Trend2), label=row.names(Z.rot.plot)) +
  theme_sleek()
#dev.off()



```


```{r}


###Diversity metrics
# 
# occAll=read.csv("occurrenceStatsForDFA.csv", header=T)
# IchSR=apply(occAll,1,sum,na.rm=T)
# 
# SWabund=read.csv('DensityForDFA.csv', header=T)
# SWabund.cut=SWabund
# IchSW=diversity(SWabund.cut, index="shannon")
# biod=data.frame(year=years.plot, IchSR, IchSW)
# write.csv(biod,'IchthyoBiod.csv', row.names=F)


plot(cor.mat$ENSO, cor.mat$trend2)

textxy(cor.mat$ENSO, cor.mat$trend2,cor.mat$X+1980)

plot(seq(1981,2013), cor.mat$trend2, type='o')
lines(seq(1981,2013), cor.mat$ENSO, type='o', col='blue')
yrs=seq(1981,2013)[-c(4,6,32)]
plot(cor.mat$ENSO,dat.z[12,])
textxy(cor.mat$ENSO[-c(4,6,32)], dat.z[12,-c(4,6,32)],yrs)

plot(cor.mat$ENSO,dat.z[22,])
textxy(cor.mat$ENSO, dat.z[22,],cor.mat$X+1980)


```

